Adversarial Deep Neural Networks Effectively Remove Nonlinear Batch Effects from Gene-Expression Data

### 2019-07-16 thesis.tex
Address comments in the manuscript

Recent progress has been made in using artificial neural networks to correct batch effects in expression data \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019}.
However, these methods are designed for a limited range of scenarios:
1) when the input data contain only two batches,
2) the batches are sufficiently large (> 100 samples in our testing), and
3) the batches are balanced.[@For \#2, does this mean the total number of samples? Or the number of samples per batch? Please clarify the wording.@]
These requirements rarely hold in applied research; for example, the bladderbatch dataset---which is used to test batch-adjustment techniques\cite{leek_bladderbatch_2017,leek_sva_2017}---has 5 batches, 57 samples, and 4 to 19 samples per batch.
Additionally, existing methods for batch correction have not been tested to verify whether nonlinear patterns remain in expression data after batch correction.
To address these limitations, we implemented a deep neural network that [@describe our method briefly, what is different about our method, and why we think our method can be applied when there are >2 batches, small datasets, and unbalanced batches.@]
Furthermore, we tested our method using machine-learning classification algorithms to quantify the extent to which nonlinear batch effects remained after adjustment.

Autoencoders are a type of neural network that encodes input data in fewer dimensions and then reconstructs the data in the original dimensions\citep{hinton_reducing_2006}[@Please verify that this is a correct reference to use here.@].
This technique can be used to reduce noise and thus refine the data representation.[@Make sure this wording is legit.@]
Typically, autoencoders are trained using an objective function that seeks to ensure that the output is as similar as possible to the input \citep{hinton_reducing_2006}.
However, neural networks are adept at finding subtle patterns in data; thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.[@Make sure this wording and the references are appropriate.@]
Recently, researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions: 1) to maximize similarity between in the input and output data and 2) to ignore any patterns associated with known confounding variables \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
Louizos, et al. have used this type of dual-objective function in combination with a variational autoencoder and successfully removed domain-based variability in credit scores, financial savings, and hospital admittance datasets \citet{louizos_variational_2015}.

@@

We trained the discriminator to predict the original batch of the autoencoder's output.
The discriminator subnetwork consists of an input layer; four fully connected hidden layers of sizes 1024, 512, 512, and 128, respectively; and an output layer sized based on the number of batches[@Please provide a little more detail about how the output layer size is calculated@].
To combat overfitting and improve training, we also added 50\%-probability dropout \cite{srivastava_dropout_2014} (which prevents overfitting by dropping a random subset of layer inputs in each training iteration) and batch normalization \cite{ioffe_batch_2015} (which helps training by smoothing out the optimization landscape \citep{santurkar_how_2018}) to each layer of the discriminator.
These additions seem to reduce overfitting in the discriminator.

@@

We initially intended to test against the SVA \citep{leek_capturing_2007} method but concluded that SVA is more suited for producing surrogate variables for further statistical research rather than removing those variables from the data.
A number of other methods for batch adjustment are available. Some of these use deep neural networks \citep{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019}, while others do not\citep{espin-perez_comparison_2018}.
Unfortunately, most of these methods lack a common interface and common assumptions that input datasets must meet.[@This won't be convincing enough to the reviewers. Be more explicit about why we couldn't use these other methods.@]

@@

We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and SVM \citep{cortes_support-vector_1995} with a radial basis kernel.
[@Say something about what hyperparameters we used. Defaults?@]

@@

[@At least in my mind, we justified using MNIST as a way to visually see how our method performs and to modify it accordingly. However, the Results don't reflect that (in my mind) because the MNIST images are blurry for Confounded, whereas they are clear for the other methods. There is some explanation below about why this might be happening with Confounded. But as an outsider reading this, I would question why we used MNIST at all because our method didn't perform particularly well. It's hard to justify using MNIST anyway because it's so different from gene-expression data. So I think we should remove it from the paper (but it was fine to use it in your thesis). Let me know if you disagree. If not, please remove it.@]

[@For simplicity, I think we should only include t-SNE graphs, not PCA. t-SNE accounts for nonlinear patterns, which is the point of our method.@]
[@I also think we should have t-SNE graphs for all three datasets.]
[@Please put in the full name for t-SNE and cite the paper for this method.@]
We used principal component analysis (PCA) and t-Stochastic... (t-SNE) plots to visualize the data before and after batch adjustment.
Before any adjustment, batches (microarray plates) in the GSE37199 dataset showed distinct expression patterns for each batch.
After adjustment with each of the three methods, the batches were less distinct (see \figurename{s} \ref{fig:pca} and \ref{fig:tsne}).
In the PCA plot, Confounded appears to maintain a similar distribution to the unadjusted data, indicating that the underlying distribution has been faithfully reproduced by the networks.
The t-SNE plot shows that the data post-adjustment by Confounded and ComBat appear to cluster less tightly by batch than the unadjusted and scale-adjusted data.
This may indicate an effective removal of nonlinear effects in both cases.
%However, previous research has shown that these plots are not completely trustworthy in representing nonlinear effects \cite{dayton_classifying_2017-1}.

>>>>

Recent progress has been made in using artificial neural networks to correct batch effects in expression data \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019}.
However, these methods are designed for a limited range of scenarios:
1) when the input data contain only two batches,
2) the batches are sufficiently large (batch sizes smaller than $n=100$ failed in our testing), and
3) the batches are balanced.[@For \#2, does this mean the total number of samples? Or the number of samples per batch? Please clarify the wording. -Is this more clear?@]
These requirements rarely hold in applied research; for example, the bladderbatch dataset---which is used to test batch-adjustment techniques\cite{leek_bladderbatch_2017,leek_sva_2017}---has 5 batches, 57 samples, and 4 to 19 samples per batch.
Additionally, existing methods for batch correction have not been tested to verify whether nonlinear patterns remain in expression data after batch correction.
To address these limitations, we implemented a deep neural network that 1) has a batch output layer that expands to fit
the number of batches, 2) works on smaller batch sizes by sampling with replacement from the training data, and 3) has
no formulas that require batches to be equal sizes. [@describe our method briefly, what is different about our method, and why we think our method can be applied when there are >2 batches, small datasets, and unbalanced batches. Is this clear enough? @]
Furthermore, we tested our method using machine-learning classification algorithms to quantify the extent to which nonlinear batch effects remained after adjustment.

Autoencoders are a type of neural network that encodes input data in fewer dimensions and then reconstructs the data in the original dimensions\citep{hinton_reducing_2006}[@Please verify that this is a correct reference to use here. - This reference looks great to me.@].
This technique can be used to reduce noise and thus refine the data representation.[@Make sure this wording is legit. - Are you referencing denoising autoencoders, or are you saying that vanilla autoencoders reduce noise? Either way, we might need a reference.@]
Typically, autoencoders are trained using an objective function that seeks to ensure that the output is as similar as possible to the input \citep{hinton_reducing_2006}.
However, neural networks are adept at finding subtle patterns in data; thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.[@Make sure this wording and the references are appropriate. - The wording and both references look great to me.@]
Recently, researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions: 1) to maximize similarity between in the input and output data and 2) to ignore any patterns associated with known confounding variables \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
Louizos, et al. have used this type of dual-objective function in combination with a variational autoencoder and successfully removed domain-based variability in credit scores, financial savings, and hospital admittance datasets \citet{louizos_variational_2015}.

@@

We trained the discriminator to predict the original batch of the autoencoder's output.
The discriminator subnetwork consists of an input layer; four fully connected hidden layers of sizes 1024, 512, 512, and 128, respectively; and an output layer sized based on the number of batches (i.e. if there are 4 distinct batches in the data, the output layer will be size 4). [@Please provide a little more detail about how the output layer size is calculated - How's this?@].
To combat overfitting and improve training, we also added 50\%-probability dropout \cite{srivastava_dropout_2014} (which prevents overfitting by dropping a random subset of layer inputs in each training iteration) and batch normalization \cite{ioffe_batch_2015} (which helps training by smoothing out the optimization landscape \citep{santurkar_how_2018}) to each layer of the discriminator.
These additions seem to reduce overfitting in the discriminator.

@@

We initially intended to test against the SVA \citep{leek_capturing_2007} method but concluded that SVA is more suited for producing surrogate variables for further statistical research rather than removing those variables from the data.
A number of other methods for batch adjustment are available. Some of these use deep neural networks \citep{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019}, while others do not\citep{espin-perez_comparison_2018}.
Unfortunately, most of these methods lack a common interface and common assumptions that input datasets must meet.[@This won't be convincing enough to the reviewers. Be more explicit about why we couldn't use these other methods. - Should I just copy the info in the background here (the "limited range of scenarios" part)?@]

@@

We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and SVM \citep{cortes_support-vector_1995} with a radial basis kernel.
In most cases, we used the default hyperparameters, with exceptions of using Random Forests with \texttt{n\_estimators=10} and using SVM with \texttt{kernel="rbf"}.
[@Say something about what hyperparameters we used. Defaults? - How's this?@]

@@

[@At least in my mind, we justified using MNIST as a way to visually see how our method performs and to modify it accordingly. However, the Results don't reflect that (in my mind) because the MNIST images are blurry for Confounded, whereas they are clear for the other methods. There is some explanation below about why this might be happening with Confounded. But as an outsider reading this, I would question why we used MNIST at all because our method didn't perform particularly well. It's hard to justify using MNIST anyway because it's so different from gene-expression data. So I think we should remove it from the paper (but it was fine to use it in your thesis). Let me know if you disagree. If not, please remove it. - We talked about maybe keeping it after training for 10k at LR=0.001 and 10k at LR=0.0001. What are your thoughts on this now?@]

[@For simplicity, I think we should only include t-SNE graphs, not PCA. t-SNE accounts for nonlinear patterns, which is the point of our method. - I like including both because it's an evidence that linear visualizations don't cut it when using ML algorithms, but I could go either way on this.@]
[@I also think we should have t-SNE graphs for all three datasets. - Do you think this should be in the main text or just in the supplement?@]
We used principal component analysis (PCA) and t-distributed Stochastic Neighbor Embedding (t-SNE) \citep{maaten_visualizing_2008} plots to visualize the data before and after batch adjustment.
Before any adjustment, batches (microarray plates) in the GSE37199 dataset showed distinct expression patterns for each batch.
After adjustment with each of the three methods, the batches were less distinct (see \figurename{s} \ref{fig:pca} and \ref{fig:tsne}).
In the PCA plot, Confounded appears to maintain a similar distribution to the unadjusted data, indicating that the underlying distribution has been faithfully reproduced by the networks.
The t-SNE plot shows that the data post-adjustment by Confounded and ComBat appear to cluster less tightly by batch than the unadjusted and scale-adjusted data.
This may indicate an effective removal of nonlinear effects in both cases.
%However, previous research has shown that these plots are not completely trustworthy in representing nonlinear effects \cite{dayton_classifying_2017-1}.
